Maximum Entropy Spectral Analysis

### 2021-02-05 paper/mesa.tex
Selected piece of theory
Changes in paper and bibliography

\usepackage{amsmath,amssymb}
\usepackage{amsfonts}
\usepackage[normalem]{ulem}
\usepackage{textcomp}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{bm}
\usepackage{afterpage}
\usepackage{graphicx}
\usepackage{psfrag}
\usepackage{mathtools}
%\usepackage{tensor}
\usepackage{layouts}
%\usepackage{DejaVuSans}
\usepackage{epstopdf}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{ragged2e}
\usepackage{blindtext}
\usepackage{graphicx}
\usepackage{siunitx}
	\sisetup{output-decimal-marker={.}}
\newtheorem{theorem}{Theorem}

@@

Jaynes Maximum Entropy principle (MAXENT) \cite{JaynesArticle}\cite{jaynes2003ptl} represents one of the most important results in Bayesian Statistics: it provides a way to uniquely assign probabilities to a phenomenon in a way that best represent our state of knowledge, while at the same time it is non committal with unavailable information. It is formulated as a variational problem in which Shannon's entropy \cite{Shannon}
\begin{equation}\label{eq:shannonEntropy}
H[p_1, \dots, p_N] = - \sum_{i = 1}^{N}p_i\log(p_i)
or, in the continuous case
H[p(x)] = - \int dx p(x)\log p(x)
is the functional to be maximized. The constraints on the variational problem are given by the prior knowledge $I$ and are usually represented as fixed values for various momenta of the probability distribution.  The most interesting result of MAXENT is obtained when a constraint on both mean and variance is considered. Consider for simplicity a vector $\vec x(t)$ with components in $\mathbb{R}^N$

@@

We assume $\vec x$ to be time dependent and real valued because we will work with time dependent single channel signals. If we put constraints in both mean $\vec \mu(t)$ and covariance matrix $C(t_i, t_j)$, MAXENT allow us to assign for $\vec x(t)$ a multivariate normal distribution \cite{gregory_2005}: 

@@

    \frac{1}{\left(2 \pi \det C\right)^{k / 2}}\exp^{\left(-\frac{1}{2}\sum_{i,j}(x_i-\mu_i) (x_j-\mu_j)C^{-1}_{ij} \right)}. 

@@

Thanks to Burg, maximum entropy method has been generalized to compute the most unbiased estimate of the spectrum for a realization of some stochastic process. 
Since it contains the same information of a power spectral density, high resolution computation of a power spectrum from a single time series $\vec x(t)$ of length $N$ is a fundamental task in signal analysis.  \\

@@

The choice of a window function is arbitrary and made by trial and error, until a satisfactory compromise between variance and resolution of the estimate of power spectral density  is reached. A high frequency resolution implies high variance and vice-versa. Another problem,is that the result depends on this choice, and this fact may raise the question of which we should consider as the "real" power spectral density. \\  

@@

@@

The choice of a window function is arbitrary and made by trial and error, until a satisfactory compromise between variance and resolution of the estimate of power spectral density  is reached. A high frequency resolution implies high variance and vice-versa. Another problem,is that the result depends on this choice, and this fact may raise the question of which we should consider as the "real" power spectral density. \\  

@@

The application of MESA is not limited to spectral estimates, but it also provides a bridge between spectral analysis and the analysis

@@

>>>>

\usepackage{amsmath,amssymb}
\usepackage{amsfonts}
\usepackage[normalem]{ulem}
\usepackage{textcomp}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{bm}
\usepackage{afterpage}
\usepackage{graphicx}
\usepackage{psfrag}
\usepackage{mathtools}
%\usepackage{tensor}
\usepackage{layouts}
%\usepackage{DejaVuSans}
\usepackage{epstopdf}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{ragged2e}
\usepackage{blindtext}
\usepackage{graphicx}
\usepackage{siunitx}
	\sisetup{output-decimal-marker={.}}
\newtheorem{theorem}{Theorem}
\usepackage{float}

@@

The Maximum Entropy principle (MAXENT) is one of the most important results in Bayesian Statistics. It provides a way to uniquely assign probabilities to a phenomenon in a way that best represent our state of knowledge, while at the same time it is non committal with unavailable information. Its domain of application turned out to be wider than expected. In fact, thanks to Burg \cite{burg1975maximum}, this method has also been applied to perform high quality computation of power spectral densities of time series. After a short introduction to Jayne's MAXENT, we will develop in detail Burg's technique of Maximum Entropy Spectral Analysis (MESA) and show that the estimate can be expressed in an analytical closed form.
for each outcome, i.e. appealing to the indifference principle (IP), is equivalent to the requirement for our experiment to be the most unpredictable. 
Consider now the opposite situation, in which probability is assigned appealing to IP and we have a different number of possible outcomes.
Intuitively, the most unpredictable experiment is the one with larger number of outcomes: probability of "getting it wrong" increases as $\frac{N -1}{N}$. So when equal probabilities are assigned, we want

@@

@@

@@

\section{Maximum Entropy Spectral Analysis} 
The computation of the power spectral density is an important component in time series analysis and characterization. Also, if the process is at least wide sense stationary, PSD is related to the autocorrelation function via a Fourier Transform.

@@

The choice of a window function is arbitrary and made by trial and error, until a satisfactory compromise between variance and resolution of the estimate of power spectral density  is reached. A high frequency resolution implies high variance and vice-versa. Another problem is that the result is very unstable with respect to the window's choice, and this fact may raise the question of which we should consider as the "real" power spectral density. \\  

@@

This approach on power spectral density estimate was developed by Burg \cite{burg1975maximum} and it is a way out from the problems raised in periodograms estimate. It requires no assumptions on the unavailable estimates for $\bar r_n$ and is consistent with the sample autocorrelation function by construction. The hard task is now to show that this problem has a solution. Remarkably, it admits an analytical solution in a closed expression for $S(f)$ 
    S(f) = \frac{P_N \Delta t}{\left(\sum_{s=0}^N a_s z^z\right)\left(\sum_{s = 0}^N a^*_s z^{-s}\right)}, 
with $\Delta t$ the sampling interval $a_0$ = 1 and the coefficients $P_N$ and $a_i (i \neq 0)$ to be determined. The explicit solution can be found in Appendix REF

\section{Autoregressive Process Analogy} 
The application of MESA is not limited to spectral estimates, but it also provides a link between spectral analysis and the analysis
Consider an autoregressive stationary process of order $p$ (AR(p)): 
    x_t - b_1 x_{t-1} - b_2 x_{t-2} \dots b_p x_{t - p} = \nu_t, 

\section{Forecasting}

\item \textbf{Optimum Bayes Decision rule (OBD)} 
The last criterion we will consider is the OBD (Rao, 1982)\cite{doi:10.1029/WR018i004p01097}. With the technique of asymptotic integration, $M$ is proposed to be the length that maximizes the posterior distribution, condition met selecting $m$ that minimize 
   \begin{equation}
        (N - m - 2) \log(P_m) + m \log(N) + \sum_{k = 0}^{m-1} \log(P_k) + \sum_{k = 1}^{m} a_k^2.
    \end{equation}
\end{itemize}
Once an optimizer is chosen to select the best recursion order $M$, we then proceed as follows:  we solve the Levinson recursion until $M_{max}$, as given in Eq.~\ref{eq:MMAx}, iterations are reached. Then, the order $M$ is selected to be the one that minimize the chosen optimizer value. \\
PARLARE QUI DELL'EARLY STOP? 
The following chapter will be devoted to the study of the statistical properties of the previous optimizers to understand which choice provides the best quality in the reproduction of some known power spectral densities.

\section{Analysis of the optimizers}

\section{Example}

\section{Comparison with Welch method}

\section{Appendix 1: MESA solution}
    \right) df.
      \frac{\delta H}{\delta \bar r_s} = \frac{1}{2Ny}\int_{-Ny}^{Ny} S(f)^{-1}e^{-\imath 2 \pi f s \Delta t } df = \lambda_s = 0 \text{ for } \vert s \vert > N. 
    S(f) = \frac{P_N \Delta t}{\left(\sum_{s=0}^N a_s z^z\right)\left(\sum_{s = 0}^N a^*_s z^{-s}\right)}. 

\begin{equation}\label{eq:OrderN}
    \begin{pmatrix}
    \bar r_0 & \bar r_{-1} & \dots &\bar r_{1 - N} & \bar r_{- N}\\
    \bar r_{1} & \bar r_0 & \dots & \bar r_{2 - N} & \bar r_{1 - N}\\ 
    \vdots & \vdots & \ddots & \vdots & \vdots \\ 
    \bar r_{N -1} & \bar r_{N-2} & \dots & \bar r_0 & \bar r_{-1}
    \\
    \bar r_{N } & \bar r_{N -1} & \dots & \bar r_1 & r_0 
    \end{pmatrix} 
    \begin{pmatrix}
    1  \\   b_1  \\  \vdots \\   b_{N - 1} \\ 0 
    \end{pmatrix} =
    \begin{pmatrix}
    P_{N -1} \\  0 \\ \vdots \\ 0 \\ \Delta_N
    \end{pmatrix} .
\end{equation}
with 
\begin{equation}
    \Delta_N = \sum_{n=0}^N \bar r_{N - n}b_n.
\end{equation}
Adding the previous expansion for both forward and backward equations, one obtains 
\begin{equation}\nonumber 
        \begin{pmatrix}
    \bar r_0 & \bar r_{-1} & \dots & \bar r_{- N}\\
    \bar r_{1} & \bar r_0 & \dots & \bar r_{1 - N}\\ 
    \vdots & \vdots & \ddots & \vdots \\ 
    \bar r_{N -1} & \bar r_{N-2} & \dots &  \bar r_{-1}
    \\
    \bar r_{N } & \bar r_{N -1} & \dots  & r_0 
    \end{pmatrix} 
    \begin{pmatrix}
    1  \\   b_1  \\  \vdots \\   b_{N - 1} \\ 0 
    \end{pmatrix} 
    + c_N 
    \begin{pmatrix}
    0 \\ b^*_{N -1} \\ \vdots  \\ b_1 \\ 1 
    \end{pmatrix}
    =
    \begin{pmatrix}
    P_{N -1} \\  0 \\ \vdots \\ 0 \\ \Delta_N
    \end{pmatrix} .
    + c_N
    \begin{pmatrix}
    \Delta^*_N \\ 0 \\\vdots \\ 0 \\ P_{N-1}
    \end{pmatrix}.
\end{equation}
With the reflection coefficient $c_N$ to be computed. Requiring the previous equation to be equivalent to \ref{eq:predictionErrorFilter}, one obtains
\begin{equation}\nonumber 
\begin{cases}
    P_{N - 1} + c_N\Delta^*_N = P_N \\
    c_N P_{N - 1} + \Delta_N = 0
\end{cases}
\end{equation}
which can be easily solved for $c_N$ and $P_N$ as 
\begin{equation}
    c_N = -\frac{\Delta_N}{P_{N-1}}; \quad \text{ and } P_N = P_{N - 1}\left(1 - \vert c_N \vert ^2 \right). 
\end{equation}
Having computed the value for the reflection coefficient, the $N$-th order forward prediction error filter is found to be 
\begin{equation}
    a_i = b_i + c_N b^*_{N - i}.
\end{equation}
To obtain the solution for the $N$-th order, one starts $\bar r_0 = P_0$ and recursively computes the new values until the desired order is computed. 
This problem is often solved numerically with Burg's Algorithm \cite{Vos}.
A python implementation of Standard Burg's Method can be found in Appendix \ref{app:Burg}

@@

The choice of a window function is arbitrary and made by trial and error, until a satisfactory compromise between variance and resolution of the estimate of power spectral density  is reached. A high frequency resolution implies high variance and vice-versa. Another problem is that the result depends on this choice, and this fact may raise the question of which we should consider as the "real" power spectral density. \\  

@@

The application of MESA is not limited to spectral estimates, but it also provides a link between spectral analysis and the analysis

@@

\chapter{Solar Spots Analysis}
\section{Introduction}
Having studied in detail MESA and understood what optimizer provides a better reproduction of spectral densities, in this chapter we provide a first and simple application of the method on the number of Sun spots as a function of time.
MESA is a natural resolution to the
problems raised and studied by Yule and Walker, since the autoregressive
component naturally arises from the data and does not require any major work.   \\
Such a study is interesting for two reasons: it is an application on a simple dataset where most of the variables can be easily taken under
control. Secondly, the study of Sun spots allow us to apply MESA for both
spectral analysis and time series forecasting. 
At first, we estimate the PSD of the time series on the whole dataset. Secondly, we will estimate the filter on a reduced dataset and use the result to forecast the time series. We will then compare the estimate with the unused, observed, data.
\section{Analysis of Sun Spots}
First attempts to give a description of the periodical behaviour of solar
activity is due to Yule and Walker \cite{doi:10.1098/rsta.1927.0007}\cite{1931RSPSA.131..518W}, who first proposed that the time series of solar spots does not have to be regarded as some
perfect sinusoidal signal with a mere superposition of fluctuations, but is
better described if a deterministic autoregressive component is considered
in the description. A description with some deterministic dependence
on the previous values of the series, allowed to reconstruct with higher precision the
time dependence of the amplitude of the process. 
In their description the choice for autoregressive order
has to be made with different trials. It seems to be clear that, at least
from this point of view, after a method for selecting the autoregressive 
order has been chosen, MESA provides a natural solution to this study. 
    \includegraphics[scale = .45]{Images/SunSpots/SunSpotsNumbers.pdf}
    \caption{Monthly number of solar spots.}
    \label{fig:sunSpotsPlot}
Fig \ref{fig:sunSpotsPlot} shows the monthly time series of Zurich Sunspots Number in a time period from $1749$ until $1983$. The periodicity in the time series is eye catching. \\
We are going to estimate the time series spectrum as a means of capturing periodicity in our dataset, MESA is applied with the FPE chosen as optimizer.  
    \includegraphics[scale = .45]{Images/SunSpots/SunSpotsPSD&ACR.pdf}
    \caption{Top panel: power spectral density for Zurich Monthly Number of Sun spots. Bottom panel: reconstructed autocorrelation function.}
    \label{fig:sunSpotsPSD&ACR}
Figure \ref{fig:sunSpotsPSD&ACR} (bottom panel) shows the the autocorrelation function. It is a slowly decaying function of time whose peaks are located at multiples of 121 months (around 10 years), reproducing the periodicity of the sample. A slowly decaying function is a sign of the process to be long range.  
Figure \ref{fig:sunSpotsPSD&ACR} (top panel) shows the frequency estimate of the Power Spectral Density obtained with MESA. The 0 frequency component is different from zero, indicating that the mean of the process is different from zero. 
There is evidence of some well defined periodicity at some frequency $f_0$, superimposed to some $f^{-\alpha}$ noise, both $f_0$ and $\alpha$ to be determined. MESA is also able to capture secondary peaks, at frequencies that are multiple of the fundamental frequency.  \\
To study it, we consider the power spectral density as a superimposition of colored noise and a Cauchy resonance. A fit of PSD background, see the top panel in Figure \ref{fig:SunSpotsFit}, shows that the decreasing component of the power spectral density can be actually treated as colored noise of the form:
    f_{noise} = (1.1 \pm .2) f^{-1.74 \pm 0.03}
    \includegraphics[scale = .45]{Images/SunSpots/SunSpotsFit.pdf}
    \caption{Power spectral density against fitting of background with colored noise (top) and with a Cauchy distribution (bottom).}
    \label{fig:SunSpotsFit}

In this way, the non periodic component can be eliminated from the spectrum and the peak can be fit. Even if not perfectly symmetrical, we chose a Cauchy distribution to fit the peak (Figure \ref{fig:SunSpotsFit}, below), parameterized as 
\begin{equation}
    C(f) = \frac{\hat{h} }{1 +\left( \frac{f - \hat f_0}{\hat \gamma}\right)^2}.
\end{equation}
The fit provided for the parameters the following values
\begin{equation}
    \nonumber 
    \hat h = (3.35 \pm 0.02)\cdot 10^5; \quad 
    \hat f_0 = (8.104 \pm 0.003) \cdot 10^{-3} \mbox{months}^{-1}; \quad 
    \hat \gamma = (5.49 \pm 0.05) \cdot 10^{-4} \mbox{months}^{-1}
\end{equation}
and the periodicity in Sun spots is found to be 
\begin{equation*}
    10.28 \pm 0.05 y_{ears}.
\end{equation*}
It is interesting to understand the meaning of the parameters by taking an inverse Fourier Transform, the autocorrelation correspondent to a pure Cauchy shaped PSD is:
\begin{equation}
    K \gamma e^{-\gamma \vert t \vert } e^{-\imath f_0 t}.
\end{equation}
$f_0$ is the natural frequency of the oscillating component for the power spectrum, while $\gamma$ is the inverse of a lifetime $\gamma = \frac{1}{\tau}$ and can be taken as a typical time scale for the autocorrelation of the periodic component, but this is not true for the whole dataset since its power spectral density is, at first approximation, the sum of the two components. Since they do not provide additional information, secondary peaks (Figure \ref{fig:sunSpotsPSD&ACR}) have been neglected.

\begin{figure}
    \centering
    \includegraphics[scale = .47]{Images/SunSpots/SunSpotsTotalFit.pdf}
    \caption{Fits to the spectrum considered with two components, one $f$ noise component and one Cauchy component for the resonance.}
    \label{fig:PSDendFit}
\end{figure}
Figure \ref{fig:PSDendFit} shows the fit obtained adding the $f$ noise component with a Cauchy resonance. The result is accurate enough to describe the main component of the power spectral density: having neglected secondary peaks does not affect the computation of the quantity of interest.

\section{Time Series Forcasting}
With MESA we have been able to compute a smooth and high quality power spectral density for the time series of Sun spots, estimating the period of solar activity to be $  10.28 \pm 0.05 y_{ears}$. This is just a first result of the application of MESA. On the other hand, the computation of the prediction error filter allows us to predict the future values for our observations as well as their credibility regions. Since we want to compare the results of the forecast with real data, we apply MESA again on our dataset, excluding the 200 last data points. We will then compare the excised data with the predictions of MESA. Having computed the prediction error filter from the data, eliminating the first value (which is one) and changing the sign to the remaining, we have an estimate of the autoregressive coefficients. In this way, The process is estimated to be an $AR(96)$ process.
    \includegraphics[scale = .47]{Images/SunSpots/SunSpotsARcoefficients.pdf}
    \caption{Estimate of the autoregressive coefficients for Sun Spots time series.}
    \label{fig:SunSpotsFilter}
Figure \ref{fig:SunSpotsFilter} shows estimate for the autoregressive coefficient of the time series. 
The behaviour of the coefficient as a function of the length is characteristic, the dependence is different from zero and positive for only few values, and becomes a noisy fluctuation around $0$ of both positive and negative components after few observations. The standard deviation $\sqrt{P_N}$ of the superimposed white noise is estimated to be $\sqrt{P_N} = 14.77$. Having computed the filter and variance for the $AR$ predictor, new values can be inferred by
    x_t = \sum_{i = 1}^{96}a_i x_{t-i} + \nu_t \text{ with } \nu_t \sim \mathbf{N}(0, \sqrt{P_N}).
In every step of the prediction, the deterministic component has to be summed to the white noise component. Since the process is linearly dependent on its previous values, this random component will propagate in the time series. For this reason, in forecasting it makes no sense to consider a realization only: even if a single prediction is an admissible realization for the process, taken alone it has no statistical meaning. \\
For this reason, to have a fair comparison between simulated and observed realization, we will simulate 3000 series of 200 points each. \\ 
Since random error propagates in the time series, we expect our prediction to be a consistent estimate of the observed time series for a number of points shorter than filter's length. In this interval, in fact, both simulation and real data are still dependent on the observed data. For predictions of length comparable with AR order or longer, our prediction is based on simulated data only and can be largely different from the observed series.
    \includegraphics[scale = .47]{Images/SunSpots/SunSpotsForecast.pdf}
    \caption{Forecasting on Sun spots time series using MESA linear predictor. Observed time series is compared with ensemble mean and 95\% Cr.}
    \label{fig:SunSpotsForecast}
Figure \ref{fig:SunSpotsForecast} shows the numerical prediction of MESA (red dashed line), computed as a mean over the ensemble, and the future values assumed by the series (black solid line) for 200 months, credibility regions at 95\% are also reported. \\ 
The vertical line at an abscissa of $96$ provides a visual separation between a time interval shorter and longer than the autoregressive order for the process. Is evident that this line separates two different regimes for the estimate. At earlier times, our inference is almost everywhere consistent with the observed data, and they are everywhere comprised in the $95\%$ regions. \\ 
At later times, the quality of inference is degraded, because the inference in this region is no more dependent on observed data. Even if the credible regions become larger there are some values laying outside them. Secondly, the observed data are no more consistent with the estimate obtained from the mean. \\
The only information we can achieve in this area is about the increasing / decreasing trend of the series. This happens because this trend is determined by the deterministic component of the process. \\ 
So, we conclude that a proper inference with MESA can be obtained for a time period shorter than the length of the filter, in that case a period of $70\sim 80 months$, i.e. around $6$ years, which is a remarkable result. \\ 
It is important to stress a fact: to obtain all the results, we assumed no knowledge on the process itself. To study power spectral density, gain quantitative information and make high quality forecast, we did not assume any physical model. All we obtained was directly contained in the data themselves and MESA extracted as much information as possible. This shows how the PSD - Autorgressive duality is powerful. With the support of Wold's theorem, given a stationary process, MESA is a tool that allow to find an approximate description of its ARMA(p,q) dual without making any assumption on the process.

\section{Conclusions}
In this section we applied MESA to a first simple dataset: the number of Sun spots. Even if simple, this dataset makes clear some important features of the method. \\ Firstly, we have been able to estimate accurately the power spectral density of the process. We have seen that, at least at the leading order, it can be treated as the sum of a colored noise component and a periodic component, described by a Cauchy resonance. The f noise component has an exponential decay with exponent $\alpha = -1.74 \pm 0.03 $ . 
The periodic component is associated to a frequency
\begin{equation}
    \nonumber 
    \hat f_0 = \left(8.104 \pm 0.003 \right) \cdot 10^{-3}. months^{-1}
\end{equation}
Secondly, we worked out with the autoregressive duality computing its AR coefficient from the prediction error filter. The AR process is found to be well approximated by an $AR(96)$ process. Having computed the coefficients, we showed how MESA can be used to  forecast over a time period shorter than the AR order $p$. This is because in that range, our estimate depends on both observed and simulated data. With no knowledge at all on the physics behind the process, we have been able to perform accurate inference and predictions with the only application of MESA. This is a very remarkable result and shows how MAXENT provide a powerful tool to make inference without superimposing any model to the dataset.

\newpage 
\chapter{Sinusoidal Signals}
\section{Introduction}
The estimation of parameters describing theoretical models is one of the key point in statistical Analysis. Considering some dataset and model, dependent on a set of parameters $\vec \theta$, the objective is to find the values for $\vec \theta$ that better describe the observation. \\
Such an estimate can only be performed if noise is correctly characterized, otherwise the inference might be imprecise.  Since data come out as a superposition of signal and noise, the characterization of noise is often made 'off-source', i.e. studied in a time window of proper length that precedes the stretch of data containing the signal. Such an analysis assumes the noise to be stationary for a period of time and is seldom justified. In this chapter, we explore a novel idea that allows to perform an "on-source" characterization of noise by the use of MESA. After a short explanation of the basic idea, we will apply this new method to a toy problem of sinusoidal signal superimposed with colored noise. 
\section{Bayesian inference}
\subsection{Bayes Theorem}
Bayes theorem is the fundamental tool for Bayesian inference. \\ 
Let $\vec \theta \in \mathbb{R}^N$ be the set of parameters we want to infer. The objective in Bayesian Analysis is the computation of the posterior distribution for the parameters given some data $D$, prior information $I$ and model $M$. \\ 
Starting from Bayes theorem, the goal is to determine the distribution $P(\vec \theta \vert D M I)$ such that 
\begin{equation}\label{eq:Bayes}
    P(\vec \theta \vert D M I) = \frac{p(D\vert \vec \theta M I)}{p(D \vert M I )} p(\vec \theta \vert M I) 
holds. \\ 
The posterior on the left hand side indicates how the parameters are distributed according to the model, prior information and the observation of some data $D$. \\ 
On the right hand side $p(\vec \theta \vert M I)$ is the prior distribution for the parameters given only model and prior information, while $p(D \vert \vec \theta M I)$ is the Likelihood. The quantity in the denominator $p(D\vert M I)$ is a purely Bayesian quantity called evidence. For sake of simplicity, since we are not going to use it explicitly, being computed as
    p(D \vert M I) = \int d\vec \theta p(D \vert \vec \theta M I) p(\vec \theta \vert M I),
we can consider it as a normalization factor. \\ It is clear that, to perform Bayesian inference,
we need to correctly determine the previous quantities. \\ 
Concerning the prior distribution, MAXENT provides a way to assign them, but there are several more methods and a huge literature about it \cite{jaynes2003ptl}. \\ 
\subsubsection{Assigning Likelihood}
Our main interest is the determination of the likelihood from the information and a Model $M$. \\
The Model $M$ defines functional relations between physical quantities, dependent on the set of parameters $\vec \theta$. Given some value $x$ for a variable, the model $M$ allows us to determine the value of a causally related variable $z$, via 
    \vec z = M(\vec x \vert \vec \theta).
The previous relations are our physical laws that we want to test and the determination of the parameters is a fundamental task to assert the validity of some Model or theory. In general, we do not observes pairs of values $(z_i, x_i)$, but we have to deal with some error that can mask the deterministic relation between the two. Out data will be "couples" of points of the form 
    D = (y_i, x_i),
where $y_i$ is determined from the model $M$ and some error $e_i$, via 
    y_i = z_i + e_i .
In general, each of the previous quantities, the observation $y_i$, the prediction from the model $z_i$ and the error $y_i$ will have their own probability distributions. The Likelihood can be obtained marginalizing the joint probability distribution over $z_i$ and $e_i$, assuming them to be independent: 
\begin{align}\nonumber 
    p_y(y_i \vert M \vec \theta I) &= \int de_i dz_i p(y_i, z_i, e_i \vert M \vec \theta I) = \int de_i dz_i p_y(y_i \vert z_i e_i M \vec \theta I) p_z(z_i \vert M \vec \theta I) p_e(e_i \vert M \vec \theta I)
Since $y_i$ is uniquely determined from the model prediction $z_i$ and the error $e_i$, its conditional distribution is 
    p(y_i \vert z_i e_i M \vec \theta I) = \delta(y_i - z_i - e_i),
so that 
\begin{align}\nonumber 
    p_y(y_i \vert M \vec \theta I) &= \int de_i dz_i \delta(y_i - z_i - e_i) p_z(z_i \vert M \vec \theta I) p_e(e_i \vert M \vec \theta I )\\ \nonumber 
    &= \int dz_i p_z(z_i \vert M \vec \theta I) p_e((y_i - z_i) \vert M \vec \theta I) 
that is the convolution between the distribution of the model and the distribution for the error. For a deterministic model, such that $z_i = m(x_i \vec \theta)$ is uniquely determined, the previous can be furhter simplified 
    p(y_i \vert M \vec \theta I) &= \int dz_i \delta\left(z_i - m(x_i \vert \vec \theta)\right) p_e\left((y_i - z_i) \vert M \vec \theta I\right) \\ \nonumber 
    & = p_e\left((y_i - m(x_i\vert \vec \theta)) \vert M \vec \theta I\right)
and for such a model the likelihood is uniquely determined from the distribution of the errors. For independent data $y_i$ and errors $e_i$, given a set of observations, the Likelihood can be rewritten as
    p(\vec y \vert M \vec \theta I) = \prod_{i}p_e\left(y_i - m(x_i \vert \vec \theta) \vert M \vec \theta I \right).
Given some deterministic model $M$, the Likelihood is uniquely determined by the distribution of the errors,
so that a full characterization of the error is a necessary step to compute the posterior for the distribution. \\ 
\newpage
\section{Power Spectral density Computation}
From the previous discussion, the role of power spectral density in parameter estimation appears to be clear: Parameter estimation can only be done if an estimate for the variance of the noise is given. \\
We briefly review how this is done for signals and propose a new 'on-source' method based on MAXENT. 
\subsection{Pure noise signals} 
Computation of power spectral density for pure noise signals of the form
    y(t) = n(t)
is the easiest task, since our observation is coincident with noise only and we don't have to extract it from the data. Also if simple, this is the very start to study even more complex data that are a superposition of signal and noise. 
The most traditional approach is based on Bayes theorem.
We can consider noise to be normally distributed with 0 mean and with some variance to be determined. To infer on the posterior distribution, a prior distribution for $\sigma$ have to be assigned
    p(\sigma \vert D I) = \frac{p(D \vert \sigma D I)}{p(D \vert I)}p(\sigma \vert I).
Firstly, we need to rewrite the previous in Frequency domain with discrete time sampling. Under these conditions, the so-called Whittle Likelihood \cite{Rover} provides a discretized version of a normal Likelihood in frequency Domain: 
    \label{eq:Whittle}
    -\frac{1}{2}\log L(\tilde n(f_i) \vert \vec S) = - \sum_{j = - N /2}^{N / 2}\left(\log(S(f_j) +  \frac{\frac{\Delta_t}{N}\vert \tilde \tilde n \vert ^2}{S(f_j)} \right), 
being $\tilde n$ the Fourier Transform of noise. In this domain, a prior on Power Spectral Density has to be assigned also. In this way, the posterior on the power spectral density can be uniquely determined as 
    p(S(f) \vert D I ) = p(\tilde n \vert S I) p(S(f) \vert I) 
This approach allows us to determine a posterior for the power spectral density and make an estimate with error bars. 
This easy example can become very complex in real cases, since the number of parameters can be arbitrarily large and a priori assignment for each of them is needed. \\
But, neglecting technicalities, such an approach is still possible. \\ 
It is interesting to see how MESA enters in this case. Having a realization of pure noise, the determination of power spectral density simply reduces to the application of MESA to the time series.
The disadvantage is that, with respect to Bayesian inference, we will only have an estimate for the power spectral density and no other quantities. The advantage is that in complicated cases, this estimate is model independent, and no assumption on noise behaviour are needed. Such a simple examples makes clear what are the main differences between MESA and model based approaches, both having their pro and cons depending on the case of study. 
\subsection{Off Source Estimate}
Even if pure noise signals are not in general our main goal, they provide a useful way to perform parameter estimation in case of data that are a superposition of both signal and noise:
d(t) = z(t) + e(t).  
In cases like this, the main objective is to compute the Posterior for the parameters $\vec \theta$ so that an estimate on their value $\vec \theta_0$ based on the observed Data can be performed. \\ 
Equation (\ref{eq:Bayes}) only formally solve this problem: to apply it, assigning likelihood is a necessary step and so posterior can be computed only after an estimate for power spectral density and noise characterization is needed. \\ 
One of the most common techniques for solving this problem is the "off source noise characterization" \cite{offsource}.
With off source studies, noise characterization is not made in the same time window where signal also lives. If signal duration is $T$, noise is studied in the $T$ seconds before signal detection. \\ 
In this way, our problem can be divided in two subsequent steps, the first one being the characterization of a pure noise signal and the second one being the estimate of parameters via Bayes theorem, that can be made with proper algorithms like MCMC.  \\
Off source characterization is an efficient and fast way to characterize noise and is very proficient when noise is stationary. 
But in the case of non stationary noise such an approach shouldn't be applied since it could bring to misleading and wrong results.

\subsection{On-source Estimate}
Our objective is to propose an 'On-source' computation to characterize noise in the same time window where signal lives, to perform a better estimate of parameters in case of non stationary noise. To avoid making assumptions on power spectral density shape, we will use MESA for its estimate and computation. This will allow us to reduce at a minimum the number of assumptions needed to perform the estimate. \\ 
We will explain our idea referring to Markov Chain Monte Carlo (MCMC) algorithm. 
\subsubsection{The Idea in a nutshell}
The conceptual simplicity of this class of algorithms is ideal to explain the idea of on source estimate without introducing blinding technicalities.  \\ 
Given some observations $D(t)$ described with some model $M(\vec \theta)$, MCMC algorithm allow to sample parameters space to find the maximum of posterior distribution for the parameters $\vec \theta$. \\
Let N be the number of recursion for the algorithm, and suppose Likelihood and prior are assigned. The structure of MCMC for posterior sampling is the following:

\begin{enumerate}
    \item Set $N = N_0$; 
    \item Set $i = 0$; 
    \item Randomly samples $\vec \theta_i$ values from prior; 
    \item Compute value for Posterior on samples $\vec \theta_i$
        P_i = P(\vec \theta = \vec \theta_i \vert D M) = p(D \vert \vec \theta_i M) * p(\vec \theta_i \vert M) 
    \item Update parameters $\theta_i$
        \vec \theta_{k} = \vec \theta_i + \vec \epsilon_i \qquad \text{with: } \vec \epsilon_i \sim D(0, \vec \sigma) 
    \item Compute $P_k$
    \item If: 
        \begin{enumerate}
            \item $P_k \geq P_i$: Set $\vec \theta_{i+1} = \vec \theta_{k}$
            \item $P_k < P_i$, Set: \\ 
            $\vec \theta_{i + 1}= \theta_k$ with probability $\frac{P_k}{P_i}$ or \\ $\vec \theta_{i+1} = \theta_i$ with probability $1 - \frac{P_k}{P_i}$
        \end{enumerate}
    \item if $i = N_0$: end
    \item increment $i \rightarrow i + 1$ 
    \item back to step 5. 
\end{enumerate}
With such an algorithm one can produce an arbitrary number of samples representative of the peak for the posterior distribution. \\ 
If the power spectral density is given, it has to be used to compute the posterior value. What happens if we don't know it? \\ 
The simplest approach is considering it as a parameter and infer it simply sampling a larger posterior distribution of the kind: 
    P(S_1, \dots, S_{N}, \theta_1, \dots, \theta_N \vert D M I). 
This would require the introduction of a prior distribution for every component of the PSD. Conceptually it is a practicable approach, but for large data set of $N$ points one should infer $N / 2$ values for the power spectral density, and complexity would increase accordingly. \\
Our objective is to propose a different approach based on MAXENT. At first, we note that at step $(6)$ the computation of the Posterior implies that we are computing the quantity: 
    d(t) - z_i(t),  
with $z_i$ the model computed on the current values $\vec \theta_i$ for the parameters. 
Collaterally, an estimate of the noise $n_i(t) = d(t) - z_i(t)$ is provided as a function of the current value for the parameters. \\ 
Having a sample of possible realization for noise, the basic idea is to compute $z_i$ power spectral density with MESA and use this estimate in the evaluation of the posterior.  \\ 
The modification needed in the algorithm to insert the computation of PSD is very simple, and can be resumed as follows: 
\begin{enumerate}
    \item Set $N = N_0$; 
    \item Set $i = 0$; 
    \item Randomly samples $\vec \theta_i$ values from prior; 
    \item Compute noise as $n_i = d - z_i$
    \item Compute power spectral density $S_i$ applying MESA on $n_i$
    \item Compute value for Posterior on samples $\vec \theta_i$ and $S_i$
        P_i = P(\vec \theta = \vec \theta_i \vert D S_i M) = p(D \vert \vec \theta_i S_i M) * p(\vec \theta_i \vert M) 
    \item Update parameters $\theta_i$
        \vec \theta_{k} = \vec \theta_i + \vec \epsilon_i \qquad \text{with: } \vec \epsilon_i \sim D(0, \vec \sigma) 
    \item Compute noise as $n_k = d - z_k$
    \item Compute power spectral density $S_k$ applying MESA on $n_k$
    \item Compute $P_k$
    \item If: 
        \begin{enumerate}
            \item $P_k \geq P_i$: Set $\vec \theta_{i+1} = \vec \theta_{k}$
            \item $P_k < P_i$, Set: \\ 
            $\vec \theta_{i + 1}= \theta_k$ with probability $\frac{P_k}{P_i}$ or \\ $\vec \theta_{i+1} = \theta_i$ with probability $1 - \frac{P_k}{P_i}$
        \end{enumerate}
